### 2023-03-02 calculus.tex
Some edits on section 2

% TODO examples of this ad hoc stuff

The key contribution of this section is the \emph{marked lambda calculus}, a formal calculus for
type error localization in bidirectional systems. Namely, the calculus provides a judgmental
framework with a unified theory for the design of typing semantics that support ill-typed programs.
\Cref{sec:calculus-calculus} describes its form and metatheory by example on a small gradually typed
lambda calculus (GTLC), extending it with binary product types in \cref{sec:calculus-products}. All
rules and theorems may be found in the supplementary appendix, alongside a complete mechanization in
the Agda proof assistant \cite{norell:thesis} (discussed in \cref{sec:calculus-agda}). With
products, we then consider how error localization is complicated by patterned let expressions in
\cref{sec:calculus-let}. Finally, \cref{sec:calculus-poly} explores how the system might be extended
to more complex typing features, such as System F-style polymorphism.

@@

We now formally introduce the marked lambda calculus through its application to the GTLC extended
with numbers and booleans. Consider the syntax for such a language, given as $\EMName$ in
\cref{fig:calculus-syntax}. The base types $\TNum$ and $\TBool$ classify number and boolean
expressions. The number literal corresponding to the mathematical number $n$ is given by $\ENumMV$,
and there is a single addition operation on numeric expressions. $\ETrue$ and $\EFalse$ correspond
to the boolean values $\textsf{true}$ and $\textsf{false}$, respectively, and
$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives a conditional structure. In addition to these forms,
$\TUnknown$ is the unknown type from gradual type theory. Finally, $\EEHole$ denotes an
\emph{expression hole}, which are used to model syntactically incomplete programs. Note, however,
that these are not critical to the calculus.

@@

Types classify expressions by a \emph{bidirectional type system}, which employs two mutually defined
judgments. \emph{Type synthesis}, written $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$, establishes that, under
the typing context $\ctx$, the expression $e$ synthesizes or locally infers \cite{Localinf} the type
$\TMV$. \emph{Type analysis}, written $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$, states that the expression
$\EMV$ may appear where an expression of type $\TMV$ is expected. We may give these typing semantics
traditionally, without considering error localization and recovery.

@@

  \item \textbf{Marking}, which \emph{marks}, i.e. type checks and localizes errors for, any unmarked
    expression into a marked expression, inserting error marks where appropriate.
\end{itemize}
%
In other words, we extend the ordinary typing semantics of a language with a secondary ``marked''
language and ``mark'' unmarked programs into marked ones. This corresponds to a type checking
process with error localization and recovery. The marked language may then serve as a foundation for
other semantic services. % such as THI.

% TODO more intuition

\Cref{fig:calculus-syntax} furthermore gives the syntax for marked expressions (as $\ECMName$),
which replicates all the forms of unmarked expressions. There are additionally several kinds of
\emph{error marks}, which represent the different type errors that may arise. Intuitively, they each
denote a different error message that might be shown in an editor or emitted by a compiler.
%
% \begin{itemize}
  % \item $\ECUnbound{x}$, the \emph{unbound variable mark}, indicates the use of an unbound variable
    % $x$, as in \cref{fig:calculus-examples-unbound}.
%
  % \item $\ECInconType{\ECMV}$, the \emph{inconsistent type mark}, indicates that the type of an
    % expression does not match an expected type.
%
  % \item $\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}$, the \emph{inconsistent branches mark}, indicates
    % that the branches of a conditional have mismatched types.
%
  % \item $\ECLamInconAsc{x}{\TMV}{\ECMV}$, the \emph{inconsistent ascription mark}, indicates that
    % the type ascription $\TMV$ is mismatches with an expected type.
%
  % \item $\ECSynNonMatchedArrow{\ECMV}$
% \end{itemize}
%
Throughout the remainder of this section, as we formulate marking for the GTLC, we will motivate and
give precise semantics for these error marks. Note that this is a non-exhaustive list; depending on
the syntactic forms and typing features found in the language, other kinds of marks may be
necessary. Indeed, in \cref{sec:calculus-poly}, we consider the kinds of marks that may be required
in an language with explicit polymorphism.

@@

Furthermore, marking should preserve the syntactic structure of programs, modulo error marks. Thus,
we define a notion of \emph{mark erasure}, whose definition is given
\cref{fig:calculus-mark-erasure}, which converts marked expressions back into unmarked expression by
removing error marks. Then, \cref{thm:calculus-marking-well-formedness} provides a
\emph{well-formedness} criterion marking.
%
\begin{theorem}[name=Marking Well-Formedness] \
  \label{thm:calculus-marking-well-formedness}
  \begin{enumerate}
    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxSynTypeM{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxAnaTypeM{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
  \end{enumerate}
  % TODO split this into two theorems?
  % TODO not really sure how to intuitively describe the part about types matching up
\end{theorem}
%
In other words, marking and then erasing error marks from the result gives the original unmarked
expression. We also ensure that marking gives well-typed marked expressions; indeed, their types
should match the types synthesized or analyzed against in marking.
% TODO more intuition for this theorem

@@

In the remainder of this subsection, we consider the different type errors that may arise in the
unmarked language while keeping in mind these two theorems; in this way, the typing semantics of the
unmarked language guide the construction of the marking process.

@@

For addition expressions, the type of both operands should be $\TNum$. To denote this in a
bidirectional system, the operands are analyzed against $\TNum$, giving the following typing rule
for unmarked addition expressions:
%
\begin{mathpar}
  \judgment{
    \ctxAnaTypeU{\ctx}{\EMV_1}{\TNum} \\
    \ctxAnaTypeU{\ctx}{\EMV_2}{\TNum}
  }{
    \ctxSynTypeU{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\TNum}
  }{USPlus}
\end{mathpar}

The marking rule mirrors \textsc{USPlus} closely. Since an expected type for each operand is known,
we shift the responsibility for any type errors to them. Hence, we recursively mark each operand in
analytic mode and rebuild the marked addition expression. The marked typing rule then mirrors
\textsc{USPlus} exactly.
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TNum} \\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TNum}
  }{
    \ctxSynFixedInto{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
  }{MKSPlus}

>>>>

The key contribution of this section is the \emph{marked lambda calculus}, a formal calculus for
bidirectional type error localization. Namely, the calculus employs bidirectional and gradual typing
to provide a judgmental framework with a unified theory for the design of typing semantics that
support ill-typed programs. \Cref{sec:calculus-calculus} describes its form and metatheory by
example on a small gradually typed lambda calculus (GTLC) \cite{Siek06a}, extending it with binary
product types in \cref{sec:calculus-products}. All rules and theorems may be found in the
supplementary appendix, alongside a complete mechanization in the Agda proof assistant
\cite{norell:thesis} (discussed in \cref{sec:calculus-agda}). With products, we then consider how
error localization is complicated by patterned let expressions in \cref{sec:calculus-let}. Finally,
\cref{sec:calculus-poly} explores how the system might be extended to more complex typing features,
such as System F-style polymorphism.

@@

We now formally introduce the marked lambda calculus through its application to the GTLC extended
with numbers and booleans. Consider the syntax for such a language, given as $\EMName$ in
\cref{fig:calculus-syntax}. The base types $\TNum$ and $\TBool$ classify number and boolean
expressions. The number literal corresponding to the mathematical number $n$ is given by $\ENumMV$,
and there is a single arithmetic operation of addition on numeric expressions. $\ETrue$ and
$\EFalse$ correspond to the boolean values $\textsf{true}$ and $\textsf{false}$, respectively, and
$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives a conditional structure.

In addition to these forms, $\TUnknown$ is the unknown type from gradual type theory \cite{Siek06a}.
Modelling the edit state of a program in development, $\EEHole$ denotes an \emph{expression hole},
used to represent syntactically incomplete portions of the program à la Hazelnut
\cite{HazelnutPOPL}. Note, however, that these are not critical to the calculus.

@@

Types classify expressions by a \emph{bidirectional type system} \cite{BidirTyping}, which employs
two mutually defined judgments. \emph{Type synthesis}, written $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$,
establishes that, under the typing context $\ctx$, the expression $e$ synthesizes or locally infers
\cite{Localinf} the type $\TMV$. \emph{Type analysis}, written $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$,
states that the expression $\EMV$ may appear where an expression of type $\TMV$ is expected. We may
give these typing semantics traditionally, without consideration for error localization and
recovery.

@@

  \item \textbf{Marking}, which \emph{marks} any unmarked expression into a marked expression,
    inserting error marks where appropriate.
\end{itemize}
%
In other words, we extend the ordinary typing semantics of a language with a secondary ``marked''
language and ``mark'' unmarked programs into marked ones. This corresponds to a type checking
process with error localization and recovery. The marked language may then serve as a foundation for
other semantic services. % TODO such as THI.

\Cref{fig:calculus-syntax} furthermore gives the syntax for marked expressions (as $\ECMName$),
which replicates all the forms of unmarked expressions. There are additionally several kinds of
\emph{error marks}, which represent the different type errors that may arise. Intuitively, they each
denote a different error message that might be shown by an editor or emitted by a compiler.

Throughout the remainder of this section, as we formulate marking for the GTLC, we will motivate and
give precise semantics for these error marks. Note that these constitute a non-exhaustive list;
depending on the syntactic forms and typing features found in a given language, other kinds of marks
may be necessary. Indeed, in \cref{sec:calculus-poly}, we consider the kinds of marks that may be
required in an language with explicit polymorphism.

@@

@@

Furthermore, since error marks are effectively annotations on top of the program, marking should
preserve syntactic structure modulo those marks. \cref{fig:calculus-mark-erasure} thus gives a
notion of \emph{mark erasure}, which converts marked expressions back into unmarked expression by
removing error marks. Then, \cref{thm:calculus-marking-well-formedness} provides a
\emph{well-formedness} criterion for marking.
%
\begin{theorem}[name=Marking Well-Formedness] \
  \label{thm:calculus-marking-well-formedness}
  \begin{enumerate}
    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxSynTypeM{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxAnaTypeM{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
  \end{enumerate}
\end{theorem}
%
In other words, marking and then erasing error marks from the result gives the original unmarked
expression. We also ensure that marking gives well-typed marked expressions; indeed, their types
should match those synthesized or analyzed against in marking.

In the remainder of this subsection, we consider the different type errors that may arise in the
unmarked language while keeping in mind these two theorems; in this way, the traditional typing
semantics of the unmarked language guide the construction of the marking process.

@@

For addition expressions, the type of both operands should be $\TNum$. To denote this in a
bidirectional system, they are analyzed against $\TNum$, giving the following typing rule for
unmarked addition expressions:
%
\begin{mathpar}
  \judgment{
    \ctxAnaTypeU{\ctx}{\EMV_1}{\TNum} \\
    \ctxAnaTypeU{\ctx}{\EMV_2}{\TNum}
  }{
    \ctxSynTypeU{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\TNum}
  }{USPlus}
\end{mathpar}

The marking rule mirrors \textsc{USPlus} closely. Since an expected type for each operand is known,
we shift the responsibility for any type errors to them. Hence, we recursively mark each operand in
analytic mode and rebuild the marked addition expression. The typing rule for marked addition
expressions then mirrors \textsc{USPlus} exactly.
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TNum} \\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TNum}
  }{
    \ctxSynFixedInto{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
  }{MKSPlus}
